Extended Data

### 2021-04-28 doc/manuscript/Dow_manuscript_Nature.Rmd
writing: discussion/ results

DOY_timing_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show leaf phenology... Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site. Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. (**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50). )" #needs attention!!!

@@

For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW. 
Specifically, greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$ (Figs. `r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).

Whereas the length of time between leaf greenup and senescence (i.e., **DEFINE**) increased in years with warmer springs (Fig. `r DOY_timing`), there was no consistent lengthening of the period of peak stem growth. 
Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, Extended Data Figures **4-6**).

**Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig. `r schematic_summary`, Extended Data Figures **4-6**).** 
Specifically, $g_{max}$ displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
while $\Delta DBH$ displayed no relationship CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`).

**To understand how annual growth increments, as registered in ring width indices (RWI), have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig. `r tree_rings`, `r EDF_map`).**
For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig. `r tree_rings`).
In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
Starting in May at SCBI and June at Harvard Forest, negative correlations of RWI to $T_{max}$ were predominant, being statistically significant in at least one growing season month for all `r n_chronologies_SCBI` species at SCBI and one of `r n_chronologies_HF` species at Harvard Forest. 
Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`).

@@

**The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (Fig. `r tree_rings`; `r EDF_map`).** 
In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only #% of records: # of # ring-porous and # or # diffuse-porous species-site combinations.
In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for #% and #% of species-site-month combinations for ring-porous and diffuse-porous species, respectively.

@@

(*length may be okay, shouldn't be lengthened*)

We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity ( `r schematic_summary`- `r DOY_timing`). Using dendroband data, we found that the peak growing season generally advanced as the critical spring windows warmed (Fig. `r DOY_timing`). Despite this shift, growth variables $L_{pgs}$. $g_{max}$ and $\Delta DBH$ were virtually unchanged when this period warmed, but $DOY_{g_{max}}$ shifted back almost in unison with $DOY_{50}$ (Fig 2). The lack of influence Spring $T_{max}$ exudes on $\Delta DBH$ found with the dendroband data agrees with patterns observed in tree cores, which show that annual tree growth depends more on peak growing season conditions than early season advancement.

Although the number of years analyzed here is limited (10 years at SCBI, 6 years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (Figs. `r schematic_summary`-`r DOY_timing`).
This indicates that the effect of $T_{max}$ on the timing of growth ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$,and $DOY_{g_{max}}$) is not subtle. 
Indeed, growth milestones advanced by 1-8 days per degree Celsius during the critical temperature window.
This is consistent with previous studies showing that ... *(find/cite some statistics on how rapidly leaf phenology or start of C sequestration advance with T--- This could be tough. It looks like that varies alot by region and species.)*

In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs. `r schematic_summary`, `r pheno_Tsensitivity`), such that a much longer record would be needed to clarify what, if any, effect spring temperatures have on maximum daily and annual growth rates. 
Fortunately, we were able to use tree-ring data to understand centennial-scale temperature sensitivity of annual growth (Fig. `r tree_rings`).
This analysis confirms that the effect of spring temperatures on annual growth is modest and not statistically significant for most species, while revealing a slightly more complex picture.
It appears that temperatures--particularly $T_{max}$--during one spring month at each site (April at SCBI, May at Harvard Forest) consistently have a weakly positive (*albiet* rarely significant) effect on annual growth, in striking contrast to the predominantly negative correlations of growth rates with $T_{max}$ in preceding and subsequent months.
This suggests that spring $T_{max}$, though it's pronounced influence on the start of woody growth and subtle influence on length of the peak growing season, has a modestly positive influence on annual tree growth, but that this influence is generally of secondary importance to other factors.

@@

@@

>>>>

DOY_timing_caption <- "**Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site. Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. (**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50). )" #needs attention!!!

@@

For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
Consistent with findings from previous studies[REFS], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively). 
Specifically, greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing 1-8 days per degree Celsius during the CTW (Figs. `r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).

Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer springs (Fig. `r DOY_timing`), there was no consistent lengthening of the period of peak stem growth. 
Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, Extended Data Figures **4-6**).

In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs. `r schematic_summary`, `r pheno_Tsensitivity`). 
Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig. `r schematic_summary`, Extended Data Figures **4-6**).
Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
while $\Delta DBH$ displayed no relationship CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth rates.

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig. `r tree_rings`, `r EDF_map`).
For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, ring width indices (RWI) rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig. `r tree_rings`).
In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
Starting in May at SCBI and June at Harvard Forest, negative correlations of RWI to $T_{max}$ were predominant, being statistically significant in at least one growing season month for all `r n_chronologies_SCBI` species at SCBI and one of `r n_chronologies_HF` species at Harvard Forest. 
Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`).

@@

The patterns observed at our focual sites were mirrored across an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (Fig. `r tree_rings`; `r EDF_map`).
In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## $^\circ$C range in April T.
Across all chronologies (focal sites included), RWI was significantly positively correlated with April $T_{max}$ for only #% of records: # of # ring-porous and # or # diffuse-porous species-site combinations.
In contrast, RWI was frequently significantly negatively correlated with $T_{max}$ during peak growing season months (May-August), with significant correlations for #% and #% of species-site-month combinations for ring-porous and diffuse-porous species, respectively.
Again, $T_{min}$ generally exerted less influence over annual growth than $T_{max}$ (`r EDF_tree_rings_TMIN`), with few significant correlations between spring $T_{min}$ and RWI.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig. `r DOY_timing`)[**REFS**] and NEE [**REFS**].
*It would be interesting to compare the magnitude of these advances (days per degree C)*
However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs.
This suggests that stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).

The concept that stem growth is not enhanced under a longer growing season was supported by our tree-ring analysis, which showed that the effect of spring temperatures on tree growth is modest and not statistically significant for most species (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).

*(KAT edits end here.)*

This analysis confirms that the effect of spring temperatures on annual growth is modest and not statistically significant for most species, while revealing a slightly more complex picture.
It appears that temperatures--particularly $T_{max}$--during one spring month at each site (April at SCBI, May at Harvard Forest) consistently have a weakly positive (*albiet* rarely significant) effect on annual growth, in striking contrast to the predominantly negative correlations of growth rates with $T_{max}$ in preceding and subsequent months.
This suggests that spring $T_{max}$, though it's pronounced influence on the start of woody growth and subtle influence on length of the peak growing season, has a modestly positive influence on annual tree growth, but that this influence is generally of secondary importance to other factors.

@@

*We have not yet re-written the methods for Nature. Nature guidelines: Full methods, typically not exceeding 3000 words. May be subdivided by bold headings.*

@@

### Leaf phenology
description here: https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description

MCD12Q2 V6 Land Cover Dynamics product (informally called the MODIS Global Vegetation Phenology product) [@friedl_mcd12q2_2019]
